In vivo targeted and deterministic single cell malignant transformation

Version 1

Abstract

Abstract

Why does a normal cell possibly harboring genetic mutations in oncogene or tumor suppressor genes becomes malignant and develop a tumor is a subject of intense debate. Various theories have been proposed but their experimental test has been hampered by the unpredictable and improbable malignant transformation of single cells. Here using an optogenetic approach we permanently turn on an oncogene (KRASG12V) in a single cell of a zebrafish brain that, only in synergy with the transient co-activation of a reprogramming factor (VENTX/NANOG/OCT4), undergoes a deterministic malignant transition and robustly and reproducibly develops within 6 days into a full-blown tumor. The controlled way in which a single cell can thus be manipulated to give rise to cancer lends support to the “ground state theory of cancer initiation” through “short-range dispersal” of the first malignant cells preceding tumor growth.

Introduction

How cancer arises from a single normal cell is still the subject of active debate, affecting intervention strategies. While many cells may harbor oncogenic mutations, only a few unpredictably end-up developing a full-blown tumor1,2,3,4. Various theories have been proposed to explain that transition5,6,7,8, but none has been tested in vivo at the single cell level. Cancer initiation is thus believed to be a rare event taking place at the level of individual cells9,10,11,12,13 arising as a result of the accumulation of genetic mutations in so called Mut-driver genes (oncogenes such as KRAS14,15, tumor suppressor genes16 such as TP53 or life-span17 genes such as TERT). Notably, mutant KRAS is the most frequent driver of several cancers18,19 : about 27% of all human cancers, 45% of colorectal and 90% of pancreatic cancers20.

Recently, it has been shown that genes involved in embryonic development, pluri/multipotency and cell reprogramming such as VENTX/NANOG and POU5/OCT4 are abnormally reactivated in late cancer stages, where acting as Epigenetic Drivers (Epi-Drivers) they empower cancer cells with Cancer Stem Cell (CSCs) features, resistance to anti-cancer therapies and potential for cancer recurrence/relapse21,22,23,24,25,26,27. Although it is evident that such Epi-Drivers confer a selective advantage to CSCs in a full-blown cancer, whether they play a role during the early phases of malignant transformation is still unknown.

Current approaches to the study of cancer use constitutive or conditional expression of Mut-driver (or Epi-driver genes28) in specific tissues, i.e. in many cells, even though only a small subset of these cells eventually leads to the growth of tumors29, often observed when the tumor already consists of many thousands of heterogeneous abnormal cells. Hence, carcinogenetic processes observed among sibling organisms30,31,32 occur with variable latency period from the onset of induction, at different locations and develop asynchronously. As a result, the initial stages of tumorigenesis are difficult to study, the state of the cell(s) of origin difficult to assess and control while its cellular environment is perturbed by the induced expression of the Mut- or Epi-driver genes. Due to the rarity of the event in vivo32, a statistically relevant single-cell tracking and characterization of the early stages of tumorigenesis has therefore never been done. This emphasises the need to predict or control the cell undergoing malignant transformation in vivo in order to pave the way for a study of the cellular and molecular events involved in the initial stages of tumorigenesis.

To address these issues, we have developed an optogenetic approach to control the expression of an oncogene in a single cell33, with the goals of: (1) measuring the probability of the malignant transformation of a single cell in a live organism in various backgrounds and (2) tracking and characterising the development of a tumor from the original cell. In the following we show that the synergy between only two factors: the oncogene kRasG12V and a reprogramming factor (Ventx, Nanog or Oct4) increases the probability of carcinogenesis from a single cell by many orders of magnitude when compared to the expression of either of these genes (or none).

Optogenetics approaches allow for the photocontrol and monitoring of the activity of biomolecules in vivo. The approach we developed uses a photoactivable analog of tamoxifen (caged cyclofen, cCYC) to control the activity of proteins fused to the ERT-receptor34,35 (a modified estrogen binding domain36) (Fig. 1A). These protein constructs are sequestered by cytoplasmic chaperones. Once cCYC is uncaged by light (with one-photon illumination at ∼375 nm or two-photon at ∼750 nm), cyclofen (CYC) is released34. It binds to the ERT-receptor and releases the fused protein (e.g. a Cre-ERT recombinase) from its complex with cytoplasmic chaperones35 (Fig. 1A).

Optogenetic set-up for single cell induction.

(A) Photo-control of a protein fused to an Estrogen Receptor (ERT) is achieved by releasing the protein from its complex with cytoplasmic chaperones (CC), upon uncaging of cCYC. (B) The transgenic zebrafish line engineered to express an oncogene (KRASG12V) upon photo-activation of a CRE-recombinase fused to ERT (CRE-ERT) (as shown in A). (C) The mRNAs of Ventx-GR and mRFP (used as a marker) are injected at the one cell stage. (D) At 1dpf the embryos are mounted in channels in an agarose gel and incubated for 45 min in cCyc on a microscope stage. (E) They are washed and illuminated at 405nm on a microscope to uncage cCyc close to the otic vesicle. A diaphragm (DIA) defines an illumination zone of ∼80μm diameter (see G). Excitation (EX), Dichroic mirrors (DM) and emission (EM) filters allow for visualization of Eos and mTFP. The cell in which KRASG12V has been induced is observed within ∼1h by the fluorescence of mTFP (see G). (F) The embryos are transferred into individual wells, incubated overnight in DEX, washed at 1dpi and monitored over the next 5 days. (G) Lateral view of zebrafish at ∼1h post-activation displays a single induced cell (top: blue spot shown by arrowhead in mTFP channel) in the illumination region (middle: Eos channel) in the vicinity of the otic vesicle (white arrow) and bottom: merger of both channels. Body axes (a: anterior; p: posterior; d; dorsal; v: ventral) are shown.

Results

We used this approach to photocontrol the activity of a Cre/loxP recombination system in a transgenic zebrafish line (Tg(actin:loxP-EOS-stop-loxP-KRASG12V-T2A-H2B-mTFP; ubi:Cre-ERT; myl7:EGFP) carrying a floxable (loxP flanked) EOS gene (coding for a green fluorescent protein) upstream from the oncogene33,34 (KRASG12V). In such a transgenic zebrafish line hereafter referred as KRASG12V line (Fig.1B, Fig.S1B) the expression of EOS can be switched to KRASG12V and H2B-mTFP (nuclear blue fluorescence) upon light-mediated uncaging of cCYC and CRE-ERT activation (Fig. 1C); mTFP fluorescence can be observed about 30min post-illumination (Fig. S1B) and is stably maintained in zebrafish (Fig. S1C). Consistent with an acquired refractory cell state and a loss of oncogenic competence30, whole body expression of the oncogene at 1dpf did not result in tumorigenesis.

To test for the possible impact of a reprogramming factor on the malignant transition, embryos from this transgenic line were injected at the one-cell stage with the mRNA of a construct consisting of a glucocorticoid receptor (GR) fused with a reprogramming factor (Ventx, Nanog or Oct4)24,37 (Fig. S3). The resulting protein (e.g. Ventx-GR) is sequestered by cytoplasmic chaperones and transiently activated upon incubation of zebrafish in Dexamethasone (Dex) (Fig. S2A). Activation of the oncogene33 at 1dpf, only if followed by transient activation of Ventx (or Nanog or Oct4) did yield reproductible hyperplasic outgrowths (Fig. S3B-E). Immuno-histo-chemistry detection of phospho-ERK activity (Fig. S4A), Hematoxylin & Eosin staining (Fig. S4B) and RT-qPCR of selected genes (Fig. S5 and S6A,B) all display features associated with tumorigenesis. None of the controls (activation of kRasG12V only, Ventx-GR only, incubation in cyclofen or dexamethasone, etc.) developed hyperplasia, but rather grew into normal zebrafish (Fig. S3).

Since both kRasG12V and Ventx have been reported to play a role in brain cancer38, we decided to activate the oncogene kRasG12V at 1dpf in a single normal cell of the brain of a transgenic zebrafish (injected with the mRNA of Ventx-GR at 1 cell stage). Activation of the oncogene in a single cell was achieved by illumination at 405nm to uncage cCyc in a small region (diameter ∼80 μm) of the brain (in the vicinity of the otic vesicle) (Fig. 1). On average the oncogene is then activated in a single cell, identified within ∼1h by the blue fluorescence of its nuclear marker (H2B-mTFP), see Fig. 1. Notice that single-cell activation of the oncogene alone does not give rise to cancer (Fig. S8A). However, if following the local expression of the oncogene (and its fluorescent marker), Ventx-GR is transiently activated the cell divides and proliferates (Fig. 2). We observed that at 1-day post induction (dpi) 50% of the induced cells had divided and expanded clonally by a factor ∼3 (Fig. 2B), while the other 50% had neither divided nor died. Surprisingly, at 3-5 dpi, we observed in all zebrafish larvae that the induced cell gave rise to progeny that display short-range dispersion (Fig. S7A, B) and then give rise to a tumor mass in the brain with local infiltration of malignant cells (Fig. 2C). Hematoxylin-Eosin (H&E) staining of the brain tissue (Fig. 2D, E) and metastases (Fig.3) further confirm the malignant state of the induced cell and its progeny.

Malignant transformation of a single cell triggering carcinogenesis in vivo.

(A) At 1dpf, a single cell in a zebrafish brain was photo-induced to express the oncogene KRASG12V, identified (white arrowhead) within ∼1h by the fluorescent H2B-mTFP. Membrane-bound mRFP is used as tracer. Transient (24h) DEX activation of Ventx is done following photoactivation. (B) At 1day post induction (1dpi), the activated cell has divided, giving rise to two cells (white arrowhead). After 3dpi the original cell expanded clonally (white arrowheads) by short-range dispersal within the brain. At 4dpi the brain has been colonized (left panel) by the progeny of the activated cell (white arrowheads) that display tumor growth as well as dispersal in the head or entering into the cardiovascular system (red arrowhead). At 5dpi, a tumor mass is formed. (C) Confocal microscopy of a larval head displaying tumors (arrowheads) and dispersal in the trunk (red arrowheads). (D) Histopathological sections of larval brain (dorsal view). Dashed lines (1 to 5) indicates sections showed in (E) Hematoxylin & Eosin (H&E) staining of brain at 5dpi of KR+VX-induced larva (depigmented) is compared to normal brain (Ctrl, melanocytes in brown). At 5dpi, the optic tectum (red arrows), the tegmentum (black arrow) and hypothalamus (red arrowhead) are infiltrated by a dysplasic tumor, progeny of the initial induced single cell. An asterisk (*) indicate the eye and a white arrow the otic vesicle. Scale bars and the body axes (a: anterior; p: posterior; d; dorsal; v: ventral) are shown. T=telencephalon; M=Mesencephalon; C= Cerebellum; Mc= Melanocytes.

In parallel with tumor mass formation in the brain (Fig. 2), we observed that some cells of the progeny re-localized to new loci far from the brain, such as the heart (Fig. 3A), the digestive tract (Fig. 3B, Fig. S6D) and the trunk (Fig. S6C). We therefore deduce that the transient activation of Ventx alters the state of a cell expressing a mutated oncogene (i.e. KRASG12V) and induces its malignant transformation in vivo, with tumorigenic potential and the capacity to generate invasive progeny.

Metastatic cells following single cell malignant transformation.

(A) A zebrafish larva, in which a single cell in the brain was photo-induced (at 1dpf) to express the oncogene KRASG12V by the blue fluorescence of the expression marker (H2b-mTFP) as in (Fig. 2), show that the progeny of the Cell of Origin of Cancer give rise to a tumor mass in the brain (white arrowhead), as well as to migrating metastatic-like cells that disseminate in the whole organism (white arrowhead), some localizing in proximity of arterial branchial arches (indicated by ζ and white arrowhead), trunk and tail fin. (B) H2b-mTFP positive cells can migrate far from the site of induction (brain) and colonize ectopic tissues located in the heart and (C) the bottom of otic vesicle (in proximity of the primary head sinus, designated by α), the digestive tract (designated by β and γ) a feature characteristic of metastatic cancer cells. Note that no anaesthetics (e.g. tricaine) or mounting media (low melting point agarose or methylcellulose) were used to block live zebrafish, during both monitoring and imaging performed on live zebrafish. An asterisk (*) indicate the eye and a white arrow the otic vesicle. Scale bars and body axes (a: anterior; p: posterior; d; dorsal; v: ventral) are indicated.

Of the larvae in which a single cell was expressing the oncogene and in which Ventx was transiently activated (KR+VX cell), all (n=15) developed tumors within 5 dpi. The frequency of tumor development is therefore F1=1. Conversely the probability for such a cell to not develop a tumor is F0=0. Due to the finite size of the sample, we estimate the probability of tumor development from a single cell to be: P1 > 80% (χ2 = 3,75; df=1).

To definitely confirm the carcinogenic nature of the KR+VX induced cells we isolated and injected a single cell from a hyperplasic tissue (identified by the blue fluorescence of its nucleus) into a naïve host zebrafish larva. This led to integration, migration and colonization of the host tissues by the progeny of the transplanted cell (Fig. 4A,B), with strong pERK activity detected in tumor masses of the host (Fig. 4B). Out of 52 transplanted host zebrafish, 31 developed tumors, a probability of tumor development (60%) consistent with previously reported efficiency of tumor cells transplantation in zebrafish39. This result definitely confirms the cancerous nature of cells expressing a mutated oncogene and exposed to the transient activation of Ventx, with subsequent alteration of the homeostasis. Notice that injection into a naïve larva of an oncogene expressing cell (identified by its nuclear blue fluorescence) which did not experience the transient activation of a reprogramming factor does not yield a tumor (Fig. S7B).

Transplantation of single cell(s) from hyperplasic tissue reveals cancer-initiating potential.

(A) Following KRASG12V plus Ventx activation, hyperplasic tissues (blue fluorescent) are observed at 6dpf, top figure. The cells of the hyperplasia were dissociated and isolated blue (KRAS expressing) cells were transplanted (≈ 1 cell per host) at 2dpf in a Nacre (mitf -/-) zebrafish line for a better tracking of the transplanted cells. The transplanted H2b-mTFP positive (blue) cell (red arrow) can be visualized as early as 3 hours post transplantation (3hpt) in the yolk of the host, close to the duct of Cuvier. White arrow indicates the head/eye (lateral view). (B) At 3 dpt the blue cell(s) from the hyperplasic tissue of the donor have colonized the host Nacre zebrafish larvae. Tumors in the brain, digestive tract and intestine, are observed and characterized by the blue fluorescence of the donor KRAS expressing cells (red arrows; n=31 out of 52 host individuals). In the bottom, immunofluorescence (IF) analysis of representative host zebrafish larvae with specific high level of phosphorylated ERK activity (pERK, red arrows) in the brain, intestine and digestive tract. (C) A high number of exogenous blue fluorescent cells are here observed to migrate in the tail (red arrows). These observations indicate that the transplanted founder cell has both migratory, colonizing behavior, as well as survival growth advantage in the host to form tumors, and thus to re-initiate carcinogenesis. Scale bars and body axes (a: anterior; p: posterior; d; dorsal; v: ventral) are shown; T=telencephalon; M=Mesencephalon; C= Cerebellum; N=notochord; VF= ventral fin.

Discussion

The surprising frequency of somatic mutations occurring in physiologically normal tissues40,41 raises the question of what combinations of events are sufficient for the malignant transformation of a single cell and the rise of the cell of origin of cancer. The robust deterministic process of single-cell cancer induction that we uncovered suggests that the aberrant reactivation of Epi-Driver genes involved in reprogramming/pluripotency (such as VENTX/NANOG, POU5/OCT4) might be relevant to the irreversible malignant transformation of a cell. Reprogramming Epi-Drivers are important regulators of cell viability, survival and proliferation in several cellular contexts, from embryonic stem cells to cancer cells22-27. Due to their capacity to modulate epigenetic memory and cell plasticity, these reprogramming factors may drive the early stages of malignant transformation in vivo once (re)activated aberrantly. They likely share some mechanism(s) with processes such as induced nuclear-reprogramming22,23,24, pluripotency maintenance and/or endogenous cell reprogramming during development37. Thus, consistent with our results, the reactivation37 of the Neural Crest (NC) Progenitor program (possibly via the stochastic expression of VENTX/NANOG and/or POU5/OCT4) has been shown in BRAF/p53 double mutant cells to yield NC-related tumors in vivo30-32.

Since cells carrying cancer-causing mutations do not deterministically developed cancers in vivo32,40, we suggest that the probability of a mutated cell to enter into malignant transformation, or to maintain its physiological functions despite mutations, correlates with the probability of the reactivation of reprogramming factors. Our results support a “Vogelgram”42 two-step model (FIG. 5) for irreversible single-cell malignant transformation in vivo, mirroring the two hits hypothesis proposed by Berenblum and Shubik43.

A two-steps “Vogelgram” model of deterministic and irreversible single-cell malignant transformation in vivo.

Within a healthy tissue, a normal cell (in light grey) acquires (step 1) genetic mutation(s) in driver oncogene(s) (Mut-Drivers such as kRASG12V). Such a mutant cell (in blue, referred as a preprocancer41 cell) can (step 2) aberrantly activate the expression of Epi-Drivers involved in pluripotency/reprogramming (e.g. VENTX/NANOG, POU5/OCT4) thus undergoing deterministic and irreversible malignant cell transformation (red cell, the Cell of Origin of Cancer); (step 3) in situ short-range dispersal of the early malignant cells and (step 4) further progression to cancer mass and to the appearance of metastatic cells. Inversely, the mutant (preprocancer) cell (in blue) can maintain its physiological functions and be eventually eliminated from the healthy tissue.

Thus, the appearance of Mut-Drivers in a normal (preprocancer41) cell40 within healthy tissue would act as a permissive but insufficient initiator for malignant cell transformation. Despite the frequency of mutational insults, which are often caused by external cues (chemical carcinogens, UV, etc…) or senescence/ageing, it is known that surveillance mechanisms like cell competition between wild-type and mutated cells safeguard tissue homeostasis and results in active elimination of mutant cells from the tissue44,45,46. Our data imply that the aberrant reactivation of Epi-Driver genes involved in reprogramming/pluripotency may promote the irreversible and deterministic malignant transformation in the cell of origin of cancer leading to carcinogenesis.

Our hypothesis is further strengthened by the observation that the variation in cancer risk among tissues can be related to the number of stem cell divisions and tissue renewal6. The aberrant reactivation (or maintenance) of Epi-Driver genes in these regenerative events, in cooperation with incident mutations, might be the key deterministic factor in the switching of a normal/healthy cell into the first cell of origin of cancer.

Furthermore, the results reported here are compatible with the recently proposed “ground state theory of cancer initiation7“. According to that theory a malignant transformation may occur in a cell harboring an oncogenic mutation upon a change of its functional state (its “ground state”). Here the transient activation of a reprogramming factor (i.e. Ventx, Nanog or Oct4) possibly synergizes with the oncogene to alter the epigenetic and functional state of the cell allowing its transformation into a tumorigenic cell.

Importantly, our observation that the progeny of the cell(s) of origin of cancer display a short-range dispersal within the tissue during the earliest phases of tumorigenesis and prior to the effective appearance of tumor mass corroborates a recent theoretical model predicting such an early dispersal during carcinogenesis to explain heterogeneity, therapeutic resistance and tumor relapse8.

Being the first experiments to predictably control the malignant transformation of a single cell in an unaltered microenvironment our approach opens a new vista on the study of “the cell of origin of cancer”47, an acknowledged enigma of cancer research. As such our results raise many questions: what mechanisms are at work in the transition from the preprocancer state to a tumorigenic state? Is the aberrant reactivation of reprogramming factors a key step in the initiation of carcinogenesis, independent of age, tissue or oncogenic mutation? Do drugs against reprogramming factors reduce the probability of carcinogenesis? Conversely, do some carcinogens (e.g. environmental pollutants, pesticides, endocrine disruptors etc…) act by reactivating those factors?

By allowing for specific and reproductible single cell malignant transformation in vivo, our optogenetic approach opens the way for a quantitative study of the initial stages of cancer at the single cell level (e.g. tracking and characterization), that will allow one to address many of these questions.

Materials and methods

Fish lines and maintenance

Zebrafish were raised and maintained in an approved Fish Facility (C75-05-32 at IBENS) on a 14–10 h light-dark diurnal cycle with standard culture methods48. Embryos collected from natural crosses were staged according to Kimmel49. The Tg(actin:loxP-EOS-stop-loxP-KRASG12V-T2A-H2B-mTFP) was generated by injecting the plasmid pT24-actin:loxP-EOS-stop-loxP-KRASG12V-T2A-H2B-mTFP, which contains the homologous cDNA sequence of KRASG12V from human, with tol2 mRNA transposase. Founder transgenic fish were identified by global expression of Eos. The Tg(ubi:Cre-ERT; myl7:EGFP) was previously described50. The double transgenic line Tg(actin:loxP-EOS-stop-loxP-KRASG12V-T2A-H2B-mTFP; ubi:Cre-ERT; myl7:EGFP) was created by crossing Tg(actin:loxP-EOS-stop-loxP-KRASG12V-T2A-H2B-mTFP) and Tg(ubi:Cre-ERT; myl7:EGFP). Founder double transgenic fish were selected by global expression of Eos and expression of EGFP in the developing heart. Zebrafish were imaged for phenotypic analysis throughout early development, from embryonic to larval stage, and then fixed (at 6dpf) with PAXgene Tissue Container Product (Qiagen) for RT-qPCR or with 4% PFA (Thermofisher) for histological (Hematoxylin & Eosine, H&E) analyses.

Dexamethasone induction

transgenic zebrafish were injected at the one cell stage with the mRNA of a gene-construct Ventx-GR (Xenopus ventx2, 450pg), Nanog-GR (mouse Nanog, 100pg) or Pou5/Oct4 (Xenopus pou5f3.1/oct91, 100pg) together with a red fluorescent expression marker mRFP (50 pg). The protein products of these constructs are sequestered by cytoplasmic chaperones and released upon incubation of the embryos in a medium containing 10 µM Dexamethasone (DEX). In single cell activation experiments, zebrafish were selected from their strong intensity of fluorescent mRFP signal for local activation of kRasG12V via uncaging of 6µM cCYC (a gift of I.Aujard and L.Jullien).

Reverse transcriptase quantitative PCR (RT-qPCR)

Zebrafish larvae were fixed (as mentioned above) at 6 dpf. Total RNAs were extracted using the RNeasy micro kit (Qiagen) according to the manufacturer’s protocols. Sample quantity and purity, reverse transcription, pre-amplification and High throughput qPCR were performed as in Zhang et al.51. Quantitative PCR was performed using the high throughput platform BioMark™ HD System and the 48.48 GE Dynamic Arrays (Fluidigm). RTqPCR measurements were done in triplicate on pooled zebrafish larvae and single zebrafish larva (for KR+VX condition), as indicated in the text. Normalization and quantification were obtained with the Δ(ΔCt) method using rpl13a as a reference gene.

Immunohistochemistry (IHC)

The zebrafish obtained in the various conditions and the stages mentioned in the text were fixed in 4% PFA overnight at 4 °C, followed by dehydration with 100% methanol at −20 °C for more than 1 day. After gradual rehydration of methanol and wash with PBS/Tween 0.1%, the embryos were incubated in a blocking solution: 1%Triton, 1% DMSO, 1% BSA and 10% sheep serum (Sigma) in PBS on a shaker for 1 h at room temperature, followed by incubation on a shaker overnight at +4°C with 1:500 anti-phospho-Erk antibody (phospho-p44/42 MAPK (Erk1/2) (Thr202/Tyr204) (D13.14.4E) XP®). Rabbit mAb detects endogenous levels of p44 and p42 MAP Kinase (Erk1 and Erk2) when dually phosphorylated at Thr202 and Tyr204 of Erk1 (Thr185 and Tyr187 of Erk2), and singly phosphorylated at Thr202 - this antibody does not cross-react with the corresponding phosphorylated residues of either JNK/SAPK or p38 MAP kinases) (ref 4370S Cell Signalling). After washing with PBS/Tween, embryos were incubated overnight at +4°C with 1/1000 secondary fluorescent conjugated antibody, Donkey anti-Rabbit Alexa Fluor 488 (Thermo Fisher A-21206). To visualize the distribution of Ventx-GR in the fish, embryos were incubated with 1:500 anti-HA antibody, tagging the HA sequence fused to Ventx-GR, as the conjugated antibody (Anti-HA, mouse IgG1, clone 16B12 Alex Fluor™ 488 conjugated, Thermofisher Ref. A21287) on a shaker overnight at 4 °C. All images were taken on a Nikon Ti microscope equipped with a Hamamatsu Orca camera, except for confocal microscopy (Fig.2C, S4B) carried out on a Leica SP5 and Zeiss LSM800 microscope.

Caged Cyclofen (cCYC) treatment and UV uncaging

To induce kRASG12V expression, 24hpf embryos were directly incubated in 6μM caged cyclofen (cCYC) for 1 hour (in the dark), briefly rinsed in E3 medium followed by 5 min. photo-activation with a ∼365 nm UV lamp. They were then transferred back to E3 medium + 10 µM DEX (to release Ventx-GR from its complex with cytoplasmic chaperones). The embryos at 1 dpi were washed 3x in E3 medium to remove DEX. For this experiment, we used a benchtop UV lamp (Fisher VL-6-L) which emits a peak wavelength at 365 nm with a FWHM (full width at half maximum) of 40 nm and delivers on the illuminated sample a typical photon flux of ∼4.3 Einstein/(s·m2).

Localized uncaging was performed by illumination for 7 min on a Nikon Ti microscope equipped with a light source peaking at 405 nm, Fig.1. The size of the uncaging region was controlled by an iris that defines a circular illumination of diameter ∼ 80 μm. After 0,5-1 h following uncaging the illuminated region of each embryo was imaged to identify and count mTFP positive cell(s). Embryos were transferred to a 12-well plate with E3 + 10 µM DEX, incubated in total darkness overnight and washed 3x with E3 at 1 dpi (to remove DEX).

Cell transplantation

Zebrafish larvae were grown until 6dpi, dissociated in dissociation medium52 (025% trypsin-EDTA plus 1/10 Collagenase 100mg/ml) 10min at 30°C, mechanically homogenized, resuspended in DMEM-10% Fetal Bovine Serum (FBS), filtered through 70µm nylon mesh (Corning Cell Strainer Ref. 431751) and resuspended in DMEM-10% FBS at 1cell/1nl concentration by using LUNA- II cell counter (Logos BioSystem).

Microscopy

Fluorescent images were taken on a Nikon Ti microscope equipped with a Hamamatsu ORCA V2+ camera and a 10X plan fluo objective. Filter setting of CFP: excitation at 438 ± 24 nm, emission at 483 ± 32 nm; mEosFP and Alexa 488: excitation at 497 ± 16 nm, emission at 535 ± 22 nm. Image analysis was done using ImageJ software.

eLife Assessment

This study provides valuable initial characterization of a verterbrate embryonic system that demonstrates aspects of an optogenetically inducible hyperplasia model. Although the evidence provided is incomplete to conclude that the system is demonstrating tumor initiation from a single cell that is metastasizing that can be quantitatively assessed, the authors propose a mechanism whereby reactivation of re-programming factors correlates with the increased likelihood of a mutant cell undergoing malignant transformation. This work will be of interest to developmental and cancer biologists mainly for the novel genetic tools described.

Reviewer #1 (Public Review):

Scerbo et al. developed an approach based on the oncogene kRasG12V and a reprogramming factor to induce deterministic and reproducible malignant transformation in a single cell. The activation of kRasG12V alone is not sufficient in their hands to initiate carcinogenesis, but when combined with the transient activation of a reprogramming factor (such as Ventx, Nanog, or Oct4), it significantly increases the probability of malignant transformation. This combination of oncogene and reprogramming factor may alter the epigenetic and functional state of the cell, leading to the development of tumors within a short period of time. The use of these two factors allows for the controlled manipulation of a single cell to study the cellular and molecular events involved in the early stages of tumorigenesis. The authors then performed allotransplantations of allegedly single fluorescent TICs in recipient larvae and found a large number of fluorescent cells in distant locations, claiming that these cells have all originated from the single transplanted TIC and migrated away. The number of fluorescent cells showed in the recipient larve just after two days is not compatible with a normal cell cycle length and more likely represents the progeny of more than one transplanted cell. The ability to migrate from the injection site should be documented by time-lapse microscopy. Then, the authors conclude that "By allowing for specific and reproducible single cell malignant transformation in vivo, their optogenetic approach opens the way for a quantitative study of the initial stages of cancer at the single cell level". However, the evidence for these claims are weak and further characterization should be performed to:

(1) show that they are actually activating the oncogene in a single cell (the magnification is too low and it is difficult to distinguish a single nucleus, labelling of the cell membrane may help to demonstrate that they are effectively activating the oncogene in, or transplanting, a single cell)

(2) the expression of the genes used as markers of tumorigenesis is performed in whole larvae, with only a few transformed cells in them. Changes should be confirmed in FACS sorted fluorescent cells

(3) the histology of the so called "tumor masses" is not showing malignant transformation, but at the most just hyperplasia. In the brain, the sections are not perfectly symmetrical and the increase of cellularity on one side of the optic tectum is compatible with this asymmetry.

(4) The number of fluorescent cells found dispersed in the larve transplanted with one single TIC after 48 hours will require a very fast cell cycle to generate over 50 cells. Do we have an idea of the cell cycle features of the transplanted TICs?

Reviewer #2 (Public Review):

Summary:

In the work by Scerbo et al, the authors aim to better understand the open question of what factors constrain cells that are genetically predisposed to form cancer (e.g. those with a potentially cancer-causing mutation like activated Ras) to only infrequently undergo this malignant transformation, with a focus on the influence of embryonic or pluripotency factors (e.g. VENTX/NANOG). Using genetically defined zebrafish models, the authors can inducibly express the KRASG12V oncogene using a combination of Cre/Lox transgenes further controlled by optogenetically inducible Cre-activated (CreER fusion that becomes active with light-induced uncaging of a tamoxifen-analogue in a targeted region of the zebrafish embryo). They further show that transient expression and activation of a pluripotency factor (e.g. Ventx fused to a GR receptor that is activated with addition of dexamethasone) must occur in the model in order for overgrowth of cells to occur. This paper describes a genetically tractable and modifiable system for studying the requirements for inducing cellular hyperplasia in a whole organism by combining overexpression of canonical genetic drivers of cancer (like Ras) with epigenetic modifiers (like specific transcription factors), which could be used to study an array of combinations and temporal relationships of these cancer drivers/modifiers.

Strengths:

The combination of Cre/lox inducible gene expression with potentially localized optogenetic induction (CreER and uncaging of tamoxifen analogues) of recombination as well as well inducible activation of a transcription factor expressed via mRNA injection (GR-fusion to the TF and dex induction) offers a flexible system for manipulating cell growth, identity, and transcriptional programs. With this system, the authors establish that Ras activation and at least transient Ventx overexpression are together required to induce a hyperproliferative phenotype in zebrafish tissues.

The ability to live image embryos over the course of days with inducible fluorophores indicating recombination events and transgene overexpression offers a tractable in vivo system for studying hyperplastic cells in the context of a whole organism.

The transplant experiments demonstrate the ability of the induced hyperplastic cells to grow upon transfer to new host.

Weaknesses:

There is minimal quantitation of key aspects of the system, most critically in the efficiency of activation of the Ras-TFP fusion (Fig 1) in, purportedly, a single cell. The authors note "On average the oncogene is then activated in a single cell, identified within ~1h by the blue fluorescence of its nuclear marker) but no additional quantitative information is provided. For a system that is aimed at "a statistically relevant single-cell

tracking and characterization of the early stages of tumorigenesis", such information seems essential.

The authors indicate that a single cell is "initiated" (Fig 2) using the laser optogenetic technique, but without definitive genetic lineage tracing, it is not possible to conclude that cells expressing TFP distant from the target site near the ear are daughter cells of the claimed single "initiated" cell. A plausible alternative explanation is 1) that the optogenetic targeting is more diffuse (i.e. some of the light of the appropriate wavelength hits other cells nearby due to reflection/diffraction), so these adjacent cells are additional independent "initiated" cells or 2) that the uncaged tamoxifen analogue can diffuse to nearby cells and allow for CreER activation and recombination. In Fig 2B, the claim is made that "the activated cell has divided, giving rise to two cells" - unless continuously imaged or genetically traced, this is unproven. In addition, it appears that Figures S3 and S4 are showing that hyperplasica can arise in many different tissues (including intestine, pancreas, and liver, S4C) with broad Ras + Ventx activation (while unclear from the text, it appears these embryos were broadly activated and were not "single cell activated using the set-up in Fig 1E? This should be clarified in the manuscript). In Fig S7 where single cell activation and potential metastasis is discussed, similar gut tissues have TFP+ cells that are called metastatic, but this seems consistent with the possibility that multiple independent sites of initiation are occurring even when focal activation is attempted.

Although the hyperplastic cells are transplantable (Fig 4), the use of the term "cells of origin of cancer" or metastatic cells should be viewed with care in the experiments showing TFP+ cells (Fig 1, 2, 3) in embryos with targeted activation for the reasons noted above.

Reviewer #3 (Public Review):

Summary:

This study employs an optogenetics approach aimed at activating oncogene (KRASG12V) expression in a single somatic cell, with a focus on following the progression of activated cell to examine tumourigenesis probabilities under altered tissue environments. The research explores the role of stemness factors (VENTX/NANOG/OCT4) in facilitating oncogenic RAS (KRASG12V)-driven malignant transformations. Although the evidence provided are incomplete, the authors propose an important mechanism whereby reactivation of re-programming factors correlates with the increased likelihood of a mutant cell undergoing malignant transformation.

Strengths:

· Innovative Use of Optogenetics: The application of optogenetics for precise activation of KRAS in a single cell is valuable to the field of cancer biology, offering an opportunity to uncover insight into cellular responses to oncogenic mutations.

· Important Observations: The findings concerning stemness factors' role in promoting oncogenic transformation are important, contributing data to the field of cancer biology.

Weaknesses:

Lack of Methodological Clarity: The manuscript lacks detailed descriptions of methodologies, making it difficult to fully evaluate the experimental design and reproducibility, rendering incomplete evidence to support the conclusion. Improving methodological transparency and data presentation will crucially strengthen the paper's contributions to understanding the complex processes of tumourigenesis.

Sub-optimal Data Presentation and Quality:

The resolution of images throughout the manuscript are too low. Images presented in Figure 2 and Figure 4 are of very low resolution. It is very hard to distinguish individual cells and in which tissue they might reside.

Lack of quantitative data and control condition data obtained from images of higher magnification limits the ability to robustly support the conclusions.

Here are some details:

· Tissue specificity of the cells express KRASG12V oncogene: In this study, the ubiquitin promoter was used to drive oncogenic KRASG12V expression. Despite this, the authors claim to activate KRAS in a single brain cell based on their localized photo-activation strategy. However, upon reviewing the methods section, the description was provided that 'Localized uncaging was performed by illumination for 7 minutes on a Nikon Ti microscope equipped with a light source peaking at 405 nm, Figure 1. The size of the uncaging region was controlled by an iris that defines a circular illumination with a diameter of approximately 80 μm.' It is surprising that an epi-fluorescent microscope with an illumination diameter of around 80μm can induce activation in a single brain cell beneath skin tissue. Additionally, given that the half-life for mTFP maturation is around 60 minutes, it is likely that more cells from a variety of different lineages could be activated, but the fluorescence would not be visible until more than 1-hour post-illumination. Authors might want to provide more evidence to support their claim on the single cell KRAS activation.

· Stability of cCYC: The manuscript does not provide information on the half-life and stability of cCYC. Understanding these properties is crucial for evaluating the system's reliability and the likelihood of leakiness, which could significantly influence the study's outcomes.

· Metastatic Dissemination claim: Typically, metastatic cancer cells migrate to and proliferate within specific niches that are conducive to outgrowth, such as the caudal hematopoietic tissue (CHT) or liver. In figure 3 A, an image showing the presence of mTFP expressing cells in both the head and tail regions of the larva, with additional positive dots located at the fin fold. This is interpreted as "metastasis" by the authors. However, the absence of a supportive cellular compartment within the fin-fold tissue makes the presence of mTFP-positive metastatic cells there particularly puzzling. This distribution raises concerns about the spatial specificity of the optogenetic activation protocol.

The unexpected locations of these signals suggest potential ectopic activation of the KRAS oncogene, which could be occurring alongside or instead of targeted activation. This issue is critical as it could affect the interpretation of whether the observed mTFP signal expansion over time is due to actual cell proliferation and infiltration, or merely a result of ectopic RAS transgene activation.

· Image Resolution Concerns: The cells depicted in Figure 3C β, which appear to be near the surface of the yolk sac and not within the digestive system as suggested in the MS, underscore the necessity for higher-resolution imaging. Without clearer images, it is challenging to ascertain the exact locations and states of these cells, thus complicating the assessment of experimental results.

· The cell transplantation experiment is lacking protocol details: The manuscript does not adequately describe the experimental protocols used for cell transplantation, particularly concerning the origin and selection of cells used for injection into individual larvae. This omission makes it difficult to evaluate the reliability and reproducibility of the results. Such as the source of transplanted cells:

• If the cells are derived from hyperplastic growths in larvae where RAS and VX (presumably VENTX) were locally activated, the manuscript fails to mention any use of fluorescence-activated cell sorting (FACS) to enrich mTFP-positive cells. Such a method would be crucial for ensuring the specificity of the cells being studied and the validity of the results.

• If the cells are obtained from whole larvae with induced RAS + VX expression, it is notable and somewhat surprising that the larvae survived up to six days post-induction (6dpi) before cells were harvested for transplantation. This survival rate and the subsequent ability to obtain single cell suspensions raise questions about the heterogeneity of the RAS + VX expressing cells that transplanted.

· Unclear Experimental Conditions in Figure S3B: The images in Figure S3B lack crucial details about the experimental conditions. It is not specified whether the activation of KRAS was targeted to specific cells or involved whole-body exposure. This information is essential for interpreting the scope and implications of the results accurately.

· Contrasting Data in Figure S3C compared to literature: The graph in Figure S3C indicates that KRAS or KRAS + DEX induction did not result in any form of hyperplastic growth. This observation starkly contrasts with previous literature where oncogenic KRAS expression in zebrafish led to significant hyper-proliferation and abnormal growth, as evidenced by studies such as those published in and Neoplasia (2018), DOI: 10.1016/j.neo.2018.10.002; Molecular Cancer (2015), DOI: 10.1186/s12943-015-0288-2; Disease Models & Mechanisms (2014) DOI: 10.1242/dmm.007831. The lack of expected hyperplasia raises questions about the experimental setup or the specific conditions under which KRAS was expressed. The authors should provide detailed descriptions of the conditions under which the experiments were conducted in Figure S3B and clarifying the reasons for the discrepancies observed in Figure S3C are crucial. The authors should discuss potential reasons for the deviation from previous reports.

Further comments:

Throughout the study, KRAS-activated cell expansion and metastasis are two key phenotypes discussed that Ventx is promoting. However, the authors did not perform any experiments to directly show that KRAS+ cells proliferate only in Ventx-activated conditions. The authors also did not show any morphological features or time-lapse videos demonstrating that KRAS+ cells are motile, even though zebrafish is an excellent model for in vivo live imaging. This seems to be a missed opportunity for providing convincing evidence to support the authors' conclusions.

There were minimal experimental details provided for the qPCR data presented in the supplementary figures S5 and S6, therefore, it is hard to evaluate result obtained.

Author response:

The following is the authors’ response to the original reviews.

First, we thank the reviewers for a thorough reading of our paper and some useful comments. A recurrent remark of the reviewers concerns the appearance of kRas-expressing cells (labelled by a nuclear blue fluorescent marker) which we attribute to the progeny of the initially induced cell. The reviewers suggest that these cells may have been obtained through activation of the Cre-recombinase in other cells by cyclofen released from light scattering, via diffusion, leakiness, etc. These remarks are perfectly reasonable from people not familiar with the cyclofen uncaging approach that we are using, but are unwarranted as we shall show below.

We have been using cyclofen uncaging with subsequent activation of a Cre-recombinase (or some other proteins) since 2010 (see ref.34, Sinha et al., Zebrafish 7, 199-204 (2010) and our 2018 review (ref.35, Zhang et al., ChemBioChem 19,1-8 (2018)). In our experiments, the embryos are incubated in the dark in 6µM caged cyclofen (cCyc) and washed in E3 medium (and transferred to a new medium with no cCyc). In these conditions, over many years we never observed activation of the recombinase, i.e. the appearance of the associated fluorescent label in cells of embryos grown in E3 medium. Hence leakiness can be ruled out (in presence of cCyc or in its absence).

Following transfer of the embryos to new E3 medium we illuminate the embryos locally with light at 405nm. In these conditions, cCyc is only partially uncaged and results in activation of Cre-recombinase in only a few cells (1,2, 3, …) within the illuminated region only, namely in the appearance of the kRas-associated nuclear blue fluorescent label in usually one cell (and sometimes in a few more). Data and statistics are now incorporated in the revised manuscript, see Fig.2A and S7. In absence of activation of a reprogramming factor these fluorescently labelled cells disappear within a few days (either via shut-down of their promotor, apoptosis or some other mechanism). The crucial point here is that we see less and not more kRas expressing cells (i.e. with nuclear blue fluorescence) in absence of VentX activation. This observation rules out activation of Cre-recombinase in other cells days after illumination due to leakiness, cyclofen released by light or diffusing from the illumination spot.

To observe many more fluorescent cells days after activation of the initial cell, one needs to transiently activate VentX-GR by overnight incubation in dexamethasone (DEX). Injecting the embryos at 1-cell stage with VentX-GR only or incubating them in DEX (without injection of VentX-GR) does not result in the appearance of more blue fluorescent cells. Following activation of VentX-GR, the fluorescent cells observed a couple of days after initiation are visualized in E3 medium (i.e. in absence of cyclofen) and are localized to the vicinity of the otic vesicle (the region where the initial cell was activated). In the revised manuscript we show images of these fluorescent cells taken a few days apart in the same embryo in which a single cell was initially activated (Fig.S8). Hence, we attribute these cells to the progeny of the activated cell. Obviously, single cell tracking via time-lapse microscopy would definitely nail down this issue and provide fascinating insight into the initial stages of tumor growth. Unfortunately, immobilization of embryos in the usual medium (e.g. MS222, tricaine) over 5-6 days to track the division and motion of single cells is not possible. We are considering some other possibilities (immobilization in bungarotoxin or via photo-activation of anionic channels), but these challenging experiments are for a future paper.

Reviewer #1 (Public Review):

The authors then performed allotransplantations of allegedly single fluorescent TICs in recipient larvae and found a large number of fluorescent cells in distant locations, claiming that these cells have all originated from the single transplanted TIC and migrated away. The number of fluorescent cells showed in the recipient larve just after two days is not compatible with a normal cell cycle length and more likely represents the progeny of more than one transplanted cell.

As mentioned in the manuscript, we measure the density of cells/nl and inject in the yolk of 2dpf Nacre embryos a volume equivalent to about 1 cell, following published protocols (S.Nicoli and M.Presta, Nat.Prot. 2,2918 (2007)). We further image the injected cell(s) by fluorescence microscopy immediately following injection, as shown in Fig.4A and Fig.S8B. We might miss a few cells but not many. With a typical cell cycle of ~10h the images of tumors in larvae at 3dpt (and not 2dpt) correspond to ~100 cells. In any case the purpose of this experiment was to show that the progeny of the initial induced cell is capable of developing into a tumor in a naïve fish, which is the operational definition of cancer that we adopted here.

The ability to migrate from the injection site should be documented by time-lapse microscopy.

As stated above our purpose here is not to study tumor formation from transplanted cell(s) but to use that assay as an operational test of cancer. Besides as mentioned earlier single cell tracking in larvea over 3-4dpt is not a trivial task.

Then, the authors conclude that "By allowing for specific and reproducible single cell malignant transformation in vivo, their optogenetic approach opens the way for a quantitative study of the initial stages of cancer at the single cell level". However, the evidence for these claims are weak and further characterization should be performed to:

(1) Show that they are actually activating the oncogene in a single cell (the magnification is too low and it is difficult to distinguish a single nucleus, labelling of the cell membrane may help to demonstrate that they are effectively activating the oncogene in, or transplanting, a single cell)

In the revised manuscript we provide larger magnification of the initial induced cell and show examples of oncogene activation in more than one cell.

(2) The expression of the genes used as markers of tumorigenesis is performed in whole larvae, with only a few transformed cells in them. Changes should be confirmed in FACS sorted fluorescent cells

When the oncogene is activated in a whole larvae all cells are fluorescent and thus FACS is of no use for cell sorting. Sorting could be done in larvae where single cells are activated , but then the efficiency of FACS is not good enough to isolate the few fluorescent cells among the many more non-fluorescent ones. We agree that the expression change of the genes used as markers of tumorigenesis is an underestimate of their true change, but our goal at this time is not to precisely measure the change in expression level, but to show that the pattern of change was different from the controls and corresponded to what is expected in tumorigenesis.

(3) The histology of the so called "tumor masses" is not showing malignant transformation, but at the most just hyperplasia.

The histology of the hyperplasic tissues show cellular proliferation with a higher density of nuclear material which is characteristic of tumors, Fig.S4C. Besides the increased expression of pERK in these tissues, Fig.S4A,B is also a hallmark of cancer.

In the brain, the sections are not perfectly symmetrical and the increase of cellularity on one side of the optic tectum is compatible with this asymmetry.

The expected T-shape formed by the sections of the tegmentum and hypothalamus are compatible with the symmetric sections shown in Fg.2D. The asymmetry in the optic tectum is a result of the hyperplasic growth.

(4) The number of fluorescent cells found dispersed in the larvae transplanted with one single TIC after 48 hours will require a very fast cell cycle to generate over 50 cells. Do we have an idea of the cell cycle features of the transplanted TICs?

As answered above, the transplanted larvae are shown at 3dpt. With a cell cycle of about 10h, a single cell can give rise to about 100 cells in that time lapse.

Reviewer #2 (Public Review):

Summary:

This paper describes a genetically tractable and modifiable system …which could be used to study an array of combinations and temporal relationships of these cancer drivers/modifiers.

We thank this referee for its positive comments. We would also like to point out that our approach provides for the first quantitative means to estimate the probability of tumorigenesis from a single cell, an estimate which is crucial in any assessment of cancer malignancy and the effectiveness of prophylactics.

Weaknesses:

There is minimal quantitation of … the efficiency of activation of the Ras-TFP fusion (Fig 1) in, purportedly, a single cell. …, such information seems essential.

We have added more images of induction of a single (or a few cells) and a plot where the probability of RAS activation in one or a few cells is specified.

The authors indicate that a single cell is "initiated" (Fig 2) using the laser optogenetic technique, but without definitive genetic lineage tracing, it is not possible to conclude that cells expressing TFP distant from the target site near the ear are daughter cells of the claimed single "initiated" cell. A plausible alternative explanation is 1) that the optogenetic targeting is more diffuse (i.e. some of the light of the appropriate wavelength hits other cells nearby due to reflection/diffraction), so these adjacent cells are additional independent "initiated" cells or 2) that the uncaged tamoxifen analogue can diffuse to nearby cells and allow for CreER activation and recombination.

We have addressed this point in our general comments to the reviewers’ remarks. The possibilities mentioned by this reviewer would result in cells expressing TFP in absence of VentX activation, which is NOT the case. Cells expressing TFP away from the initial site are observed DAYS after activation of the oncogene (and TFP) in a single cell and ONLY upon activation of VentX.

In Fig 2B, the claim is made that "the activated cell has divided, giving rise to two cells" - unless continuously imaged or genetically traced, this is unproven.

We have addressed this remark previously. Tracking of larvae over many days is not possible with the usual protocol using tricaine to immobilize the larvae. Nonetheless, in the revised version we present images of an embryo imaged at various times post activation (1hpi, 3dpi, 7dpi) where proliferation and metastasis of the cells can be observed. We are pursuing other alternatives for time-lapse microscopy over many days, since besides convincing the sceptics, a single cell tracking experiment (possibly coupled with in-situ spatial transcriptomics) will shed a new and fascinating light on the initial stages of tumor growth.

In addition, it appears that Figures S3 and S4 are showing that hyperplasia can arise in many different tissues (including intestine, pancreas, and liver, S4C) with broad Ras + Ventx activation …. This should be clarified in the manuscript).

This is true and has been clarified in the new version.

In Fig S7 where single cell activation and potential metastasis is discussed, similar gut tissues have TFP+ cells that are called metastatic, but this seems consistent with the possibility that multiple independent sites of initiation are occurring even when focal activation is attempted.

As mentioned previously this is ruled out by the fact that these cells are observed days after cyclofen uncaging (and TFP activation) and IF AND ONLY IF VentX was activated during the first dpi.

Although the hyperplastic cells are transplantable (Fig 4), the use of the term "cells of origin of cancer" or metastatic cells should be viewed with care in the experiments showing TFP+ cells (Fig 1, 2, 3) in embryos with targeted activation for the reasons noted above.

The purpose of this transplantation experiment was to show that cell in which both kRas and VentX have been activated possess the capacity to metastasize and develop a tumor mass when transplanted in a naïve zebrafish. This - to the best of our knowledge - is the operational definition of a malignant tumor. Notice also that transplantation of kRAS only activated cells (i.e. without subsequent activation of VentX) does NOT yield tumors, rather the transplanted cell disappears after a few days, see Fig.S10.

Reviewer #3 (Public Review):

Summary:

This study employs an optogenetics approach … to examine tumorigenesis probabilities under altered tissue environments.

We thank this reviewer for this remark, since we believe that the probability to assess the probability of tumorigenesis from a single cell is probably the most significant contribution of this work.

Weaknesses:

Lack of Methodological Clarity: The manuscript lacks detailed descriptions of methodologies,

We have included additional detail of our methodology and statistical analyses in the revised manuscript.

Sub-optimal Data Presentation and Quality:

Lack of quantitative data and control condition data obtained from images of higher magnification limits the ability to robustly support the conclusions.

We have included more images at higher magnification and quantitative data to support the main report of targeted single cell induction.

Here are some details:

Authors might want to provide more evidence to support their claim on the single cell KRAS activation.

More images and a data on activation of single or few cells in the illumination field are provided as well as statistical analysis of cell induction.

Stability of cCYC: The manuscript does not provide information on the half-life and stability of cCYC. Understanding these properties is crucial for evaluating the system's reliability and the likelihood of leakiness, which could significantly influence the study's outcomes.

We have been using the cCyc system for about 14 years. We refer the reader to our previous papers and reviews on this methodology. Briefly, cCyc is stable when not illuminated with light around 375nm. Typically, we incubate our embryos in the dark for about 1h before washing, transferring them into E3 medium and illuminating them. Assessing the leakiness of the system is easy as expression of a fluorescent marker is permanently turned on. We have observed none in the conditions of our experiment or in previous works.

Metastatic Dissemination claim: However, the absence of a supportive cellular compartment within the fin-fold tissue makes the presence of mTFP-positive metastatic cells there particularly puzzling. This distribution raises concerns about the spatial specificity of the optogenetic activation protocol … The unexpected locations of these signals suggest potential ectopic activation of the KRAS oncogene,

We have addressed this remark in the introduction and above. Specifically, metastatic and proliferative mTFP-positive cells are observed IF AND ONLY IF VentX is also activated concomitant with activation of kRAS in a single cell. No proliferative cells are observed in absence of VentX activation, or in presence of VentX or Dex alone, or if kRAS has not been activated by cyclofen uncaging.

Image Resolution Concerns: The cells depicted in Figure 3C β, which appear to be near the surface of the yolk sac and not within the digestive system as suggested in the MS, underscore the necessity for higher-resolution imaging. Without clearer images, it is challenging to ascertain the exact locations and states of these cells, thus complicating the assessment of experimental results.

Better images are provided in the revised version.

The cell transplantation experiment is lacking protocol details:

Details are provided. We have followed regular protocols for transplantation: S.Nicoli and M.Presta, Nat.Prot. 2,2918 (2007).

If the cells are obtained from whole larvae with induced RAS + VX expression, it is notable and somewhat surprising that the larvae survived up to six days post-induction (6dpi) before cells were harvested for transplantation. This survival rate and the subsequent ability to obtain single cell suspensions raise questions about the heterogeneity of the RAS + VX expressing cells that transplanted.

From Fig.S4D, about 50% of the embryos survive at 6dpi. Though an interesting question by itself we have not (yet) addressed the important issue of the heterogeneity of the outgrowth obtained from a single cell. Our purpose here was just to show that cells in which both kRAS and VentX have been activated possess the capacity to metastasize and develop a tumor mass when transplanted in a naïve zebrafish. This - to the best of our knowledge - is the operational definition of a malignant tumor.

Unclear Experimental Conditions in Figure S3B: …It is not specified whether the activation of KRAS was targeted to specific cells or involved whole-body exposure.

This was whole body (global) illumination and is specified in the revised version.

Contrasting Data in Figure S3C compared to literature: The graph in Figure S3C indicates that KRAS or KRAS + DEX induction did not result in any form of hyperplastic growth. The authors should provide detailed descriptions of the conditions under which the experiments were conducted in Figure S3B and clarifying the reasons for the discrepancies observed in Figure S3C are crucial. The authors should discuss potential reasons for the deviation from previous reports.

This discrepancy is discussed in the revised version. First the previous reports consider the development of tumors within 3-4 weeks which we have not studied in detail. Second, the expression of the oncogene in these reports might be stronger than in ours. Third, the stochastic and random appearance of tumors in these reports suggest that some other mechanism (transient stress-induced reprogramming?) might have activated the oncogene in the initial cell.

Further comments:

Throughout the study, KRAS-activated cell expansion and metastasis are two key phenotypes discussed that Ventx is promoting. However, the authors did not perform any experiments to directly show that KRAS+ cells proliferate only in Ventx-activated conditions.

Yes, we did. See Fig. S1 and compare with Fig.S3B, or Fig.S10A in comparison with Fig.2A,B.

The authors also did not show any morphological features or time-lapse videos demonstrating that KRAS+ cells are motile, even though zebrafish is an excellent model for in vivo live imaging. This seems to be a missed opportunity for providing convincing evidence to support the authors' conclusions.

Performing time-lapse microscopy on larvae over many (4-5) days is not possible with the regular tricaine protocol for immobilization. We are definitely planning such experiments, but they will require some other protocol, perhaps using bungarotoxin or some optogenetic inhibitory channels.

There were minimal experimental details provided for the qPCR data presented in the supplementary figures S5 and S6, therefore, it is hard to evaluate result obtained.

More details are given in the revised version.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Abstract: what is the definition of tumors that they are using? I never heard of a full-blown tumor that develops in less than 6 days from a single cell!

This is indeed surprising! We are using an operational definition of a tumor: if cells from an hyperplasic tissue can metastasize and outgrow when transplanted in a naïve zebrafish, then it is a tumor.

Introduction: The claim that this is the first report of the induction of oncogene expression in a single cell in zebrafish is wrong as there are other reports (PMID: 27810924, PMID: 30061297)

These other approaches are invasive (electroporation and transplantation). We have added non-invasive in the revised version.

Figure 2: The quality of these images is too low to visualize the infiltration that they talk about, the sections are not perfectly coronal and the asymmetric distribution of cells may be confused with an infiltration.

We have addressed this question above.

Results, page 5: how do we know that these are metastatic cells? there could have been spurious activation in other locations, you need to prove that these cells moved from one place to the other and that they are of the same cell type as the primary tumor

We have addressed this question extensively in the introduction and in our answers to the reviewers. We have also added a figure showing cell proliferation in the same embryos at various time post induction. Time-lapse microscopy studies of tumor initiation and growth over many days are planned, but will be the subject of an other paper.

Figure 3: not clear why they did not use anaesthetic or mounting media to take pictures of the transplanted fish

We tried to minimally stress the larvae that are already in a perilous condition…

Results, page 6: Not clear why the authors used KRAS v12 as an oncogene and uncaged its expression in the brain, as KRAS is not a common oncogene for brain tumors.

There are reports of kRASG12V tumors in zebrafish brain (doi: 10.1186/s12943-015-0288-2)

It is not clear what is the mechanism of Ventx -driven oncogenesis? What changes in gene expression, cell function etc are induced by Ventx in the cells that express KRASv12? The qPCR analysis performed is done on whole larvae and an analysis on single TICs and their progeny should be done following FACS sorting of fluorescent cells.

FACS sorting of a single TIC (and its progeny) among many thousand cells in the embryo is not possible. The analysis on whole larvae provides an underestimate of the changes in gene expression following activation of kRAS and VentX. We are looking for spatial transcriptomics as a better approach of the changes in gene expression induced in single TICs and their progeny, but that is beyond the scope of this paper.

Nuclear staining is necessary to make sure that only 1 cell was transplanted. How is it possible that we get more than 50 cells from a single transplanted cell in less than 48 hours? What is the length of the cell cycle of these transformed cells?

Nuclear staining is not necessary as the transplanted cell is fluorescent. Thus we can see how many cells are transplanted. With a cell-cycle of about 10h in 3dpt, a single cell will have generated as many as 100 cells.

Reviewer #2 (Recommendations For The Authors):

Minor grammatical change - hyperplasic more commonly called hyperplastic.

Reviewer #3 (Recommendations For The Authors):

Provide Detailed Methodologies: Clearly describe all experimental protocols used, particularly those for cell transplantation and photo-activation techniques. Detailed protocols will aid in replicating your findings and enhancing the manuscript's credibility.

Done.

Provide High-Resolution Imaging data: To substantiate the claims about cell location and behaviour, provide high-resolution images where individual cells and their specific tissue contexts are clearly visible.

Greater magnification images provided.

Quantitative Data: Incorporate quantitative analyses to strengthen the findings, particularly in experiments where cell proliferation and activation are key outcomes.

Done.

Verify Single Cell Activation: Offer additional evidence or experimental validation to support the claim that KRASG12V activation is confined to single cells, considering the limitations mentioned about the photo-activation setup.

Discussion, figures and statistical analysis added in manuscript.

Discuss Stability and Leakage of cCYC: Provide data on the stability and half-life of cCYC to assess the likelihood of system leakiness, which could influence the interpretation of your results.

Reference to our previous papers and reviews added.

Clarify Metastatic Claims: Discuss the unexpected presence of mTFP-positive cells in nontraditional metastatic sites, like the fin fold, and consider additional experiments to verify whether these are cases of ectopic activation or true metastasis.

Discussion added in manuscript

Utilize time-lapse live imaging to visually document the motility and behaviour of KRAS+ cells over time, leveraging the strengths of the zebrafish model.

Definitely interesting, but non trivial to conduct over many days and subject for a future paper.

Address Discrepancies in KRAS Activation Effects from literature: Specifically, discuss why your findings on KRAS-induced hyperplasia differ from existing literature. Consider whether experimental conditions or KRAS expression levels might have contributed to these differences.

Discussion added in revised version

Version 2

Abstract

Abstract

Why does a normal cell possibly harboring genetic mutations in oncogene or tumor suppressor genes becomes malignant and develops a tumor is a subject of intense debate. Various theories have been proposed but their experimental test has been hampered by the unpredictable and improbable malignant transformation of single cells. Here using an optogenetic approach we permanently turn on an oncogene (KRASG12V) in a single cell of a zebrafish brain that, only in synergy with the transient co-activation of a reprogramming factor (VENTX/NANOG/OCT4), undergoes a deterministic malignant transition and robustly and reproducibly develops within 6 days into a full-blown tumor. The controlled way in which a single cell can thus be manipulated to give rise to cancer lends support to the “ground state theory of cancer initiation” through “short-range dispersal” of the first malignant cells preceding tumor growth.

Introduction

How cancer arises from a single normal cell is still the subject of active debate, affecting intervention strategies. While many cells may harbor oncogenic mutations, only a few unpredictably end-up developing a full-blown tumor1,2,3,4. Various theories have been proposed to explain that transition5,6,7,8, but none has been tested in vivo at the single cell level. Cancer initiation is thus believed to be a rare event taking place at the level of individual cells9,10,11,12,13 arising as a result of the accumulation of genetic mutations in so called Mut-driver genes (oncogenes such as KRAS14,15, tumor suppressor genes16 such as TP53 or life-span17 genes such as TERT). Notably, mutant KRAS is the most frequent driver of several cancers18,19: about 27% of all human cancers, 45% of colorectal and 90% of pancreatic cancers20.

Recently, it has been shown that genes involved in embryonic development, pluri/multipotency and cell reprogramming such as VENTX/NANOG and POU5/OCT4 are abnormally reactivated in late cancer stages, where acting as Epigenetic Drivers (Epi-Drivers) they empower cancer cells with Cancer Stem Cell (CSCs) features, resistance to anti-cancer therapies and potential for cancer recurrence/relapse21,22,23,24,25,26,27. Although it is evident that such Epi-Drivers confer a selective advantage to CSCs in a full-blown cancer, whether they play a role during the early phases of malignant transformation is still unknown.

Current approaches to the study of cancer use constitutive or conditional expression of Mut-driver (or Epi-driver genes28) in specific tissues, i.e. in many cells, even though only a small subset of these cells eventually leads to the growth of tumors29, often observed when the tumor already consists of many thousands of heterogeneous abnormal cells. Hence, carcinogenetic processes observed among sibling organisms30,31,32 occur with variable latency period from the onset of induction, at different locations and develop asynchronously. As a result, the initial stages of tumorigenesis are difficult to study, the state of the cell(s) of origin difficult to assess and control while its cellular environment is perturbed by the induced expression of the Mut-or Epi-driver genes. Due to the rarity of the event in vivo32, a statistically relevant single-cell tracking and characterization of the early stages of tumorigenesis has therefore never been done. This emphasises the need to predict or control the cell undergoing malignant transformation in vivo in order to pave the way for a study of the cellular and molecular events involved in the initial stages of tumorigenesis.

To address these issues, we have developed an optogenetic approach to control the expression of an oncogene in a single cell33, with the goals of: (1) measuring the probability of the malignant transformation of a single cell in a live organism in various backgrounds and (2) tracking and characterising the development of a tumor from the original cell. In the following we show that the synergy between only two factors: the oncogene kRasG12V and a reprogramming factor (Ventx, Nanog or Oct4) increases the probability of carcinogenesis from a single cell by many orders of magnitude when compared to the expression of either of these genes (or none).

Optogenetics approaches allow for the photocontrol and monitoring of the activity of biomolecules in vivo. The approach we developed uses a photoactivable analog of tamoxifen (caged cyclofen, cCYC) to control the activity of proteins fused to the ERT-receptor34,35 (a modified estrogen binding domain36) (Fig. 1A). These protein constructs are sequestered by cytoplasmic chaperones. Once cCYC is uncaged by light (with one-photon illumination at ∼375 nm or two-photon at ∼750 nm), cyclofen (CYC) is released34. It binds to the ERT-receptor and releases the fused protein (e.g. a Cre-ERT recombinase) from its complex with cytoplasmic chaperones35 (Fig. 1A).

Optogenetic set-up for single cell induction.

(A) Photo-control of a protein fused to an Estrogen Receptor (ERT) is achieved by releasing the protein from its complex with cytoplasmic chaperones (CC), upon uncaging of cCYC. (B) The transgenic zebrafish line engineered to express an oncogene (KRASG12V) upon photo-activation of a CRE-recombinase fused to ERT (CRE-ERT) (as shown in A). (C) The mRNAs of Ventx-GR and mRFP (used as a marker) are injected at the one cell stage. (D) At 1dpf the embryos are mounted in channels in an agarose gel and incubated for 45 min in cCyc on a microscope stage. (E) They are washed and illuminated at 405nm on a microscope to uncage cCyc close to the otic vesicle. A diaphragm (DIA) defines an illumination zone of ∼80μm diameter (see G). Excitation (EX), Dichroic mirrors (DM) and emission (EM) filters allow for visualization of Eos and mTFP. The cell in which KRASG12V has been induced is observed within ∼1h by the fluorescence of mTFP (see G). (F) The embryos are transferred into individual wells, incubated overnight in DEX, washed at 1dpi and monitored over the next 5 days. (G) Lateral view of zebrafish at ∼1h post-activation displays a single induced cell (top: blue spot shown by arrowhead in mTFP channel) in the illumination region (middle: Eos channel) in the vicinity of the otic vesicle (white arrow) and bottom: merger of both channels. Body axes (a: anterior; p: posterior; d; dorsal; v: ventral) are shown.

Results

We used this approach to photocontrol the activity of a Cre/loxP recombination system in a transgenic zebrafish line (Tg(actin:loxP-EOS-stop-loxP-KRASG12V-T2A-H2B-mTFP; ubi:Cre-ERT; myl7:EGFP) carrying a floxable (loxP flanked) EOS gene (coding for a green fluorescent protein) upstream from the oncogene33,34 (KRASG12V). In such a transgenic zebrafish line hereafter referred as KRASG12V line (Fig.1B, Fig.S1B) the expression of EOS can be switched to KRASG12V and H2B-mTFP (nuclear blue fluorescence) upon 1h incubation in cCYC, followed by washing and UV (365 nm) cCYC uncaging which triggers CRE-ERT activation (Fig. 1C); mTFP fluorescence can be observed about 30min post-illumination (Fig. S1B) and is stably maintained in zebrafish (Fig.1G, Fig. S1C). Consistent with an acquired refractory cell state and a loss of oncogenic competence30, whole body expression of the oncogene at 1dpf did not result in tumorigenesis. Notice that consistent with previous studies33–35, where cCYC stability has been validated both in vitro (by chemical analyses), and in vivo (with fluorescent reporters of uncaging), in absence of illumination no spontaneous uncaging of cCYC and thus activation of the oncogene (i.e. blue fluorescent cells) is observed. Notice also that once cCYC is uncaged embryos are maintained in E3 medium with no cyclofen and thus no activation of the oncogene.

To test for the possible impact of a reprogramming factor on the malignant transition, embryos from this transgenic line were injected at the one-cell stage with the mRNA of a construct consisting of a glucocorticoid receptor (GR) fused with a reprogramming factor (Ventx, Nanog or Oct4)24,37 (Fig.S2). The resulting protein (e.g. Ventx-GR) is sequestered by cytoplasmic chaperones and transiently activated upon incubation of zebrafish in Dexamethasone (DEX) (Fig. S2A). Activation of the oncogene33 at 1dpf, if and only if followed by transient activation of Ventx (or Nanog or Oct4) did yield reproductible hyperplasic outgrowths in many different tissues (including brain, intestine, pancreas and liver; Fig. S3B-E). Immuno-histo-chemistry detection of phospho-ERK activity (Fig. S4A), Hematoxylin & Eosin staining (Fig. S4B) and RT-qPCR of selected genes (Figs. S5; S6A,B) all display features associated with tumorigenesis. None of the controls (activation of kRasG12V only, Ventx-GR only, incubation in cyclofen or dexamethasone, etc.) developed hyperplasia, but rather grew into normal zebrafish (Figs. S1B,C; S3A). These controls imply that the observed tumor is only a result of the joint activation of kRasG12V (permanently) and a reprogramming factor (transiently).

Since both kRasG12V and Ventx have been reported to play a role in brain cancer38, we decided to activate the oncogene kRasG12V at 1dpf in a single normal cell of the brain of a transgenic zebrafish (injected with the mRNA of Ventx-GR at 1 cell stage). Activation of the oncogene in a single cell was achieved by 1h incubation in cCYC, followed by washing and illumination at 405nm to uncage cCyc in a small area (diameter ∼80 μm) of the brain (in the vicinity of the otic vesicle) (Fig. 1G). Notice that following uncaging the embryos are maintained in a cyclofen free E3 medium.

Subsequent to cCYC uncaging, with probability ∼50% the oncogene is activated in a single cell of the illuminated area (activation statistics is shown in Fig.2A), identified within ∼1h by the blue fluorescence of its nuclear marker (H2B-mTFP), see Figs. 1G, 2A, S7. H2B-mTFP positive cells are only observed at the site of illumination (i.e. brain), attesting to the precision of our optogenetic method (Fig. 2A). This observation specifies whether one, two or more cells were activated (Fig. S7) within ∼1h post-illumination. Notice that activation of the oncogene alone does not give rise to cancer (the activated cell usually disappears within a few days, Fig. S10A), which implies that the proliferation of kRasG12V positive (mTFP+, blue fluorescent) cells - reported next - is not due to late activation of the oncogene (and its fluorescent marker).

Malignant transformation of a single cell triggering carcinogenesis in vivo.

(A) At 1dpf, a single cell in a zebrafish brain was photo-induced to express the oncogene KRASG12V, identified (white arrowhead) within ∼1h by the blue fluorescent H2B-mTFP. Membrane-bound mRFP is used as tracer. Transient (24h) DEX activation of Ventx is done following photoactivation. The probability p of inducing one or more (blue fluorescent) cells is shown on the right panel, together with the Poisson distribution (red curve) expected for the independent induction of k cells (error bars are statistical errors on the mean: where N is the total number of observed embryos). (B) At 1day post induction (1dpi), the activated cell may have divided (middle panel) giving rise to two mTFP+ (blue fluorescent) cells (white arrowhead) or may have not divided (left panel). The probability of observing k blue fluorescent cells at 1dpi is shown on the right panel. (C) At 3dpi the original cell expanded clonally (white arrowheads) by short-range dispersal within the brain. At 4dpi the brain has been colonized (middle panel) by the progeny of the activated cell that display tumor growth as well as dispersal in the head or entering into the cardiovascular system (red arrowhead). At 5dpi (right panel), a tumor mass is formed. (D) Confocal microscopy of a larval head displaying tumors (white arrowheads) and dispersal in the trunk (red arrowheads). (E) Histopathological sections of larval brain (dorsal view). Dashed lines (1 to 5) indicates sections showed in (F) Hematoxylin & Eosin (H&E) staining of brain at 5dpi of KR+VX-induced larva (depigmented) is compared to normal brain (Ctrl, melanocytes in brown). At 5dpi, the optic tectum (red arrows), the tegmentum (black arrow) and hypothalamus (red arrowhead) are infiltrated by a dysplasic tumor, progeny of the initial induced single cell. An asterisk (*) indicate the eye and a white arrow the otic vesicle. Scale bars and the body axes (a: anterior; p: posterior; d; dorsal; v: ventral) are shown. T=telencephalon; M=Mesencephalon; C= Cerebellum; Mc= Melanocytes (yellow arrows).

Indeed, if and only if following the local expression of the oncogene, Ventx-GR is transiently activated (by 24h incubation in dexamethasone) does the cell divide and proliferate (Figs. 2B-C, S8, S9). We observed that at 1-day post induction (dpi) ∼50% of the induced cells had divided and expanded clonally by a factor ∼3 (Fig. 2B). In the other 50% of cases, the activated single cell had neither divided nor died (Fig. 2B). Surprisingly, at 3-5 dpi, we observed in all zebrafish larvae (n=30) that the induced cell(s) gave rise to progeny that display short-range dispersion (Fig. S9A, B) and then give rise to a tumor mass in the brain with local infiltration of malignant cells (Fig. 2C) Hematoxylin-Eosin (H&E) staining of the brain tissue (Fig. 2D, E) and metastases (Fig.3) further confirm the malignant state of the induced cell and its progeny.

Metastatic cells following single cell malignant transformation.

(A) A zebrafish larva, in which a single cell in the brain was photo-induced (at 1dpf) to express the oncogene KRASG12V by the blue fluorescence of the expression marker (H2b-mTFP) as in (Fig. 2), show that the progeny of the Cell of Origin of Cancer give rise to a tumor mass in the brain (white arrowhead), as well as to migrating metastatic-like cells that disseminate in the whole organism (white arrowhead), some localizing in proximity of arterial branchial arches (indicated by ζ and white arrowhead), trunk and tail fin. (B) H2b-mTFP positive cells can migrate far from the site of induction (brain) and colonize ectopic tissues located in the heart and (C) the bottom of otic vesicle (in proximity of the primary head sinus, designated by α), the digestive tract (designated by β and γ) a feature characteristic of metastatic cancer cells. Note that no anaesthetics (e.g. tricaine) or mounting media (low melting point agarose or methylcellulose) were used to block live zebrafish, during both monitoring and imaging performed on live zebrafish. An asterisk (*) indicate the eye and a white arrow the otic vesicle. Scale bars and body axes (a: anterior; p: posterior; d; dorsal; v: ventral) are indicated.

In parallel with tumor mass formation in the brain (Fig. 2), we observed that some cells of the progeny re-localized to new loci far from the brain, such as the heart (Fig. 3A, the digestive tract (Fig. 3B, Fig. S6D) and the trunk (Fig. S6C, Fig.S8B). We therefore deduce that the transient activation of Ventx alters the state of a cell expressing a mutated oncogene (i.e. KRASG12V) and induces its malignant transformation in vivo, with tumorigenic potential and the capacity to generate invasive progeny. Notice that metastatic cells must be the progeny of the initial (blue fluorescent) cell as the larvae are incubated in embryo medium with no cCYC. Furthermore, absent VentX activation we do not observe proliferation and metastatic behaviour, rather the initially induced cell disappears, see Fig.S10A.

Of the larvae in which a few (1-6) cells were expressing the oncogene and in which Ventx was transiently activated (KR+VX cell), all (N = 30) developed tumors within 5 dpi. The frequency of tumor development is therefore F1=1. Conversely the probability for such a cell to not develop a tumor is F0=0. Due to the finite size of the sample, we estimate the probability of tumor development from a single cell to be: P1 > 80% (χ2 = 3,75; df=1).

To definitely confirm the carcinogenic nature of the KR+VX induced cells we isolated and injected a single cell from a hyperplasic tissue (identified by the blue fluorescence of its nucleus) into a naïve host zebrafish larva. This led to integration, migration and colonization of the host tissues by the progeny of the transplanted cell (Fig. 4A,B), with strong pERK activity detected in tumor masses of the host (Fig. 4B). Out of 52 transplanted host zebrafish, 31 developed tumors, a probability of tumor development (60%) consistent with previously reported efficiency of tumor cells transplantation in zebrafish39. The capacity of the progeny of the induced cells to metastasize and develop a tumor mass when transplanted in a naïve zebrafish is - to the best of our knowledge - the operational definition of a malignant tumor. This result definitely confirms the cancerous nature of cells expressing a mutated oncogene and exposed to the transient activation of Ventx, with subsequent alteration of the homeostasis. Notice that injection into a naïve larva of an oncogene expressing cell (identified by its nuclear blue fluorescence) which did not experience the transient activation of a reprogramming factor does not yield a tumor and disappears within a few days (Fig. S10B).

Transplantation of single cell(s) from hyperplasic tissue reveals cancer-initiating potential.

(A) Following KRASG12V plus Ventx activation, hyperplasic tissues (blue fluorescent) are observed at 6dpf, top figure. The cells of the hyperplasia were dissociated and isolated blue (KRAS expressing) cells were transplanted (≈ 1 cell per host) at 2dpf in a Nacre (mitf -/-) zebrafish line for a better tracking of the transplanted cells. The transplanted H2b-mTFP positive (blue) cell (red arrow) can be visualized as early as 3 hours post transplantation (3hpt) in the yolk of the host, close to the duct of Cuvier. White arrow indicates the head/eye (lateral view). (B) At 3 dpt the blue cell(s) from the hyperplasic tissue of the donor have colonized the host Nacre zebrafish larvae. Tumors in the brain, digestive tract and intestine, are observed and characterized by the blue fluorescence of the donor KRAS expressing cells (red arrows; n=31 out of 52 host individuals). In the bottom, immunofluorescence (IF) analysis of representative host zebrafish larvae with specific high level of phosphorylated ERK activity (pERK, red arrows) in the brain, intestine and digestive tract. (C) A high number of exogenous blue fluorescent cells are here observed to migrate in the tail (red arrows). These observations indicate that the transplanted founder cell has both migratory, colonizing behavior, as well as survival growth advantage in the host to form tumors, and thus to re-initiate carcinogenesis. Scale bars and body axes (a: anterior; p: posterior; d; dorsal; v: ventral) are shown; T=telencephalon; M=Mesencephalon; C= Cerebellum; N=notochord; VF= ventral fin.

Discussion

The surprising frequency of somatic mutations occurring in physiologically normal tissues40,41 raises the question of what combinations of events are sufficient for the malignant transformation of a single cell and the rise of the cell of origin of cancer. The robust deterministic process of single-cell cancer induction that we uncovered suggests that the aberrant reactivation of Epi-Driver genes involved in reprogramming/pluripotency (such as VENTX/NANOG, POU5/OCT4) might be relevant to the irreversible malignant transformation of a cell. Reprogramming Epi-Drivers are important regulators of cell viability, survival and proliferation in several cellular contexts, from embryonic stem cells to cancer cells22–27. Due to their capacity to modulate epigenetic memory and cell plasticity, these reprogramming factors may drive the early stages of malignant transformation in vivo once (re)activated aberrantly. They likely share some mechanism(s) with processes such as induced nuclear-reprogramming22,23,24, pluripotency maintenance and/or endogenous cell reprogramming during development37. Thus, consistent with our results, the reactivation37 of the Neural Crest (NC) Progenitor program (possibly via the stochastic expression of VENTX/NANOG and/or POU5/OCT4) has been shown in BRAF/p53 double mutant cells to yield NC-related tumors in vivo30–32.We surmise that activation of Epi-Driver genes in NC-related tumors might confer predictable tumorigenesis.

Since cells carrying cancer-causing mutations do not deterministically developed cancers in vivo32,40, we suggest that the probability of a mutated cell to enter into malignant transformation, or to maintain its physiological functions despite mutations, correlates with the probability of the reactivation of reprogramming factors. Our results support a “Vogelgram”42 two-step model (FIG. 5) for irreversible single-cell malignant transformation in vivo, mirroring the two hits hypothesis proposed by Berenblum and Shubik43.

A two-steps “Vogelgram” model of deterministic and irreversible single-cell malignant transformation in vivo.

Within a healthy tissue, a normal cell (in light grey) acquires (step 1) genetic mutation(s) in driver oncogene(s) (Mut-Drivers such as kRASG12V). Such a mutant cell (in blue, referred as a preprocancer41 cell) can (step 2) aberrantly activate the expression of Epi-Drivers involved in pluripotency/reprogramming (e.g. VENTX/NANOG, POU5/OCT4) thus undergoing deterministic and irreversible malignant cell transformation (red cell, the Cell of Origin of Cancer); (step 3) in situ short-range dispersal of the early malignant cells and (step 4) further progression to cancer mass and to the appearance of metastatic cells. Inversely, the mutant (preprocancer) cell (in blue) can maintain its physiological functions and be eventually eliminated from the healthy tissue.

Thus, the appearance of Mut-Drivers in a normal (preprocancer41) cell40 within healthy tissue would act as a permissive but insufficient initiator for malignant cell transformation. Despite the frequency of mutational insults, which are often caused by external cues (chemical carcinogens, UV, etc…) or senescence/ageing, it is known that surveillance mechanisms like cell competition between wild-type and mutated cells safeguard tissue homeostasis and results in active elimination of mutant cells from the tissue44,45,46. Our data imply that the aberrant reactivation of Epi-Driver genes involved in reprogramming/pluripotency may promote the irreversible and deterministic malignant transformation in the cell of origin of cancer leading to carcinogenesis.

Our hypothesis is further strengthened by the observation that the variation in cancer risk among tissues can be related to the number of stem cell divisions and tissue renewal6. The aberrant reactivation (or maintenance) of Epi-Driver genes in these regenerative events, in cooperation with incident mutations, might be the key deterministic factor in the switching of a normal/healthy cell into the first cell of origin of cancer.

Furthermore, the results reported here are compatible with the recently proposed “ground state theory of cancer initiation7”. According to that theory a malignant transformation may occur in a cell harboring an oncogenic mutation upon a change of its functional state (its “ground state”). Here the transient activation of a reprogramming factor (i.e. Ventx, Nanog or Oct4) possibly synergizes with the oncogene to alter the epigenetic and functional state of the cell allowing its transformation into a tumorigenic cell.

This may also explain the apparent discrepancy between the present deterministic and rapid tumor development upon activation in a single cell of both kRasG12V and a reprogramming factor and the reports of stochastic and random tumor development over weeks following activation of kRasG12V in a whole tissue (brain47, pancreas48 or intestine49). In the latter case, transformed cells may have been triggered to undergo a malignant transformation by the stochastic activation of a reprogramming factor/Epi-Driver, which is consistent with the recent finding that cancer initiation might occurs by transient perturbations of epigenetic regulators50.

Importantly, our observation that the progeny of the cell(s) of origin of cancer display a short-range dispersal within the tissue during the earliest phases of tumorigenesis and prior to the effective appearance of tumor mass corroborates a recent theoretical model predicting such an early dispersal during carcinogenesis to explain heterogeneity, therapeutic resistance and tumor relapse8.

Being the first experiments to predictably and non-invasively control the malignant transformation of a single cell in an unaltered microenvironment our approach opens a new vista on the study of “the cell of origin of cancer”51, an acknowledged enigma of cancer research. As such our results raise many questions: what mechanisms are at work in the transition from the preprocancer state to a tumorigenic state? Is the aberrant reactivation of reprogramming factors a key step in the initiation of carcinogenesis, independent of age, tissue or oncogenic mutation? Do drugs against reprogramming factors reduce the probability of carcinogenesis? Conversely, do some carcinogens (e.g. environmental pollutants, pesticides, endocrine disruptors etc…) act by reactivating those factors?

By allowing for specific and reproductible single cell malignant transformation in vivo, our optogenetic approach opens the way for a quantitative study of the initial stages of cancer at the single cell level (e.g. tracking and characterization), that will allow one to address many of these questions.

Materials & methods

Fish lines and maintenance

Zebrafish were raised and maintained in an approved Fish Facility (C75- 05-32 at IBENS) on a 14–10 h light-dark diurnal cycle with standard culture methods52. Embryos collected from natural crosses were staged according to Kimmel53. The Tg(actin:loxP-EOS-stop-loxP- KRASG12V-T2A-H2B-mTFP) was generated by injecting the plasmid pT24-actin:loxP-EOS-stop-loxP- KRASG12V-T2A-H2B-mTFP, which contains the homologous cDNA sequence of KRASG12V from human, with tol2 mRNA transposase. Founder transgenic fish were identified by global expression of Eos. The Tg(ubi:Cre-ERT; myl7:EGFP) was previously described54. The double transgenic line Tg(actin:loxP-EOS-stop-loxP-KRASG12V-T2A-H2B-mTFP; ubi:Cre-ERT; myl7:EGFP) was created by crossing Tg(actin:loxP-EOS-stop-loxP-KRASG12V-T2A-H2B-mTFP) and Tg(ubi:Cre-ERT; myl7:EGFP). Founder double transgenic fish were selected by global expression of Eos and expression of EGFP in the developing heart. Zebrafish were imaged for phenotypic analysis throughout early development, from embryonic to larval stage, and then fixed (at 6dpf) with PAXgene Tissue Container Product (Qiagen) for RT-qPCR or with 4% PFA (Thermofisher) for histological (Hematoxylin & Eosine, H&E) analyses.

Caged Cyclofen (cCYC) treatment and UV uncaging

To induce kRASG12V expression, 24hpf embryos were incubated in 6μM caged cyclofen (cCYC) for 1 hour (in the dark), rinsed in E3 medium followed by 5 min. photo-activation with a ∼365 nm UV lamp (measured intensity: 4mW/cm2). They were then transferred back to E3 medium (with no cCYC) + 10 µM DEX (to release Ventx-GR from its complex with cytoplasmic chaperones). The embryos at 1 dpi were washed 3x in E3 medium to remove DEX. From thereon embryos are exposed neither to cCYC, nor to DEX. For this experiment, we used a benchtop UV lamp (Fisher VL-6-L) which emits a peak wavelength at 365 nm with a FWHM (full width at half maximum) of 40 nm and delivers on the illuminated sample a typical photon flux of ∼1,25 10-4 Einstein/(s·m2).

Localized uncaging was performed by illumination for 7 min on a Nikon Ti microscope equipped with a light source peaking at 405 nm, Fig.1. The size of the uncaging region was controlled by an iris that defines a circular illumination of diameter ∼ 80 μm. After 0,5-1 h following uncaging the illuminated region of each embryo was imaged to identify and count mTFP positive cell(s) (cells in which the oncogene and its reporter fluorophore was activated). In about 50% of cases one cell was thus photo-activated, the other ∼50% corresponded to more than one activated cell, see Fig.2A. Embryos were transferred to a 12-well plate with E3 + 10 µM DEX, incubated in total darkness overnight and washed 3x with E3 at 1 dpi (to remove DEX).

Dexamethasone induction

transgenic zebrafish were injected at the one cell stage with the mRNA of a gene-construct Ventx-GR (Xenopus ventx2, 450pg), Nanog-GR (mouse Nanog, 100pg) or Pou5/Oct4 (Xenopus pou5f3.1/oct91, 100pg) together with a red fluorescent expression marker mRFP (50 pg). The protein products of these constructs are sequestered by cytoplasmic chaperones and released upon incubation of the embryos in a medium containing 10 µM Dexamethasone (DEX). In single cell activation experiments, zebrafish were selected from their strong intensity of fluorescent mRFP signal for local activation of kRasG12V via uncaging of 6µM cCYC (a gift of I.Aujard and L.Jullien).

Reverse transcriptase quantitative PCR (RT-qPCR)

Zebrafish larvae were fixed (as mentioned above) at 6 dpf. Total RNAs were extracted using the RNeasy micro kit (Qiagen) according to the manufacturer’s protocols. Sample quantity and purity, reverse transcription, pre-amplification and High throughput qPCR were performed as in Zhang et al.55. Specifically:

Reverse transcription

cDNA synthesis was performed using Reverse Transcription Master Mix from Standard Biotools according to the manufacturer’s protocol with random primers in a final volume of 5 μL containing 40 ng total RNA using a Nexus thermocycler (Eppendorf). cDNA samples were diluted by adding 20 μL of low TE buffer [10 mM Tris; 0.1 mM EDTA; pH = 8.0 (TEKNOVA)] and stored at - 20°C.

Specific target amplification

1.25 μL of each diluted cDNA was used for multiplex pre-amplification with Standard Biotools® PreAmp Master Mix at 18 cycles. In a total volume of 5 μL. the reaction contained 1 μL of pre-amplification mastermix, 1.25 μL of cDNA, 1.25 μL of pooled TaqMan® Gene Expression assays (Life Technologies, ThermoFisher) with a final concentration of each assay of 180 nM (0.2X) and 1 μL of PCR water. The cDNA samples were subjected to pre-amplification following the temperature protocol: 95°C for 2 min, followed by 18 cycles at 95°C for 15 s and 60°C for 4 min. The pre-amplified cDNA were diluted 5X by adding 20 μL of low TE buffer (TEKNOVA) and stored at - 20°C before qPCR.

High-throughput real time PCR

quantitative PCR was performed using the high-throughput platform BioMark™ HD System and the 48.48 GE Dynamic Arrays (Standard Biotools). 6 μL of sample master mix (SMM) consisted of 1.8 μL of 5X diluted pre-amplified cDNA, 0.3 μL of 20X GE Sample Loading Reagent (Standard Biotools) and 3 μL of TaqMan® Gene Expression PCR Master Mix (Life Technologies, ThermoFisher). Each 6 μL assay master mix (AMM) consisted of 3 μL of TaqMan® Gene Expression assay 20X (Life Technologies) and 3 μL of 2X Assay Loading Reagent (Standard Biotools). 5 μL of each SMM and each AMM premixes were added to the dedicated wells. The samples and assays were mixed inside the chip using MX-FC controller (Standard Biotools). Thermal conditions for qPCR were: 25°C for 30 min and 70°C for 60 min for thermal mix; 50°C for 2 min and 95°C for 10 min for hot start; 40 cycles at 95°C for 15 s and 60°C for 1 min. Data were processed by automatic threshold for each assay, with linear derivative baseline correction using BioMark Real-Time PCR Analysis Software 4.0.1 (Standard Biotools). The quality threshold was set at the default setting of 0.65.

RT-qPCR measurements were done in triplicate on pooled zebrafish larvae and single zebrafish larva (for KR+VX condition), as indicated in the text. Normalization and quantification were obtained with the ΔΔCt method using rpl13a as a reference gene. The relative expression of the genes R under the different conditions analysed was calculated as follows using the method described by Livak&Schmittgen56: R = 2-ΔΔCt where ΔΔCt = (Ct gene of interest - Ct reference gene) test condition - (Ct gene of interest – Ct reference gene) control condition

Immunohistochemistry (IHC)

The zebrafish obtained in the various conditions and stages mentioned in the text were fixed in 4% PFA overnight at 4 °C, followed by dehydration with 100% methanol at −20 °C for more than 1 day. After gradual rehydration of methanol and wash with PBS/Tween 0.1%, the embryos were incubated in a blocking solution: 1%Triton, 1% DMSO, 1% BSA and 10% sheep serum (Sigma) in PBS on a shaker for 1 h at room temperature, followed by incubation on a shaker overnight at +4°C with 1:500 anti-phospho-Erk antibody (phospho-p44/42 MAPK (Erk1/2) (Thr202/Tyr204) (D13.14.4E) XP®). Rabbit mAb detects endogenous levels of p44 and p42 MAP Kinase (Erk1 and Erk2) when dually phosphorylated at Thr202 and Tyr204 of Erk1 (Thr185 and Tyr187 of Erk2), and singly phosphorylated at Thr202 - this antibody does not cross-react with the corresponding phosphorylated residues of either JNK/SAPK or p38 MAP kinases) (ref 4370S Cell Signalling). After washing with PBS/Tween, embryos were incubated overnight at +4°C with 1/1000 secondary fluorescent conjugated antibody, Donkey anti-Rabbit Alexa Fluor 488 (Thermo Fisher A-21206). To visualize the distribution of Ventx-GR in the fish, embryos were incubated with 1:500 anti-HA antibody, tagging the HA sequence fused to Ventx-GR, as the conjugated antibody (Anti-HA, mouse IgG1, clone 16B12 Alex FluorTM 488 conjugated, Thermofisher Ref. A21287) on a shaker overnight at 4 °C. All images were taken on a Nikon Ti microscope equipped with a Hamamatsu Orca camera, except for confocal microscopy (Fig.2C, S4B) carried out on a Leica SP5 and Zeiss LSM800 microscope.

Cell transplantation

Transplantation experiments followed published protocols57.Specifically, zebrafish larvae were grown until 6dpi, dissociated in dissociation medium58 (025% trypsin-EDTA plus 1/10 Collagenase 100mg/ml) 10min at 30°C, mechanically homogenized, resuspended in DMEM-10% Fetal Bovine Serum (FBS), filtered through 70µm nylon mesh (Corning Cell Strainer Ref. 431751) and resuspended in DMEM-10% FBS at 1cell/1nl concentration by using LUNA-II cell counter (Logos BioSystem), and injected (∼1 nl) at the level of the yolk sac circulation valley (close to the site of the ducts of Cuvier) of host nacre zebrafish at 2 days post-fertilization (2dpf) as shown in Fig.4.

Microscopy

Fluorescent images were taken on a Nikon Ti microscope equipped with a Hamamatsu ORCA V2+ camera and a 10X plan fluo objective. Filter setting of CFP: excitation at 438 ± 24 nm, emission at 483 ± 32 nm; mEosFP and Alexa 488: excitation at 497 ± 16 nm, emission at 535 ± 22 nm. Image analysis was done using ImageJ software.